---
title: "Untitled"
author: "Steve Simon"
date: "November 12, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

## Here's a data set to illustrate overfitting

```{r read}
fn <- "http://www.statsci.org/data/oz/firearms.txt"
d0 <- read.table(file=fn, header=TRUE, as.is=TRUE)
d0
cat("Source:", fn)
d1 <- d0[-15, ]
d2 <- data.frame(Year=seq(d0$Year[1], d0$Year[15], by=0.125))
n1 <- which(d2$Year %in% 1983:1996)
n2 <- which(d2$Year==1997)
```

## Here's some code to display the data and the predictions.

First, display the data.

```{r function1, echo=TRUE}
draw_raw_data <- function() {
  # Eliminate top margin
  par(mar=c(4.1, 4.1, 0.1, 0.6))
  # Draw the basic plotting frame without any points
  plot(d0$Year, d0$Rate, type="n", ylim=c(2,5),
    xlab="Year", ylab="Death rate per 100,000")
  # Show the values used in the regression model in black
  text(d0$Year[-15], d0$Rate[-15], d0$Rate[-15], col="black", cex=0.75)
  # Show the value not used in gray
  text(d0$Year[ 15], d0$Rate[ 15], d0$Rate[ 15], col="gray", cex=0.75)
}
```

## Here's some code to display the data and the predictions.

Next, code for displaying predictions

```{r function2, echo=TRUE}
draw_predictions <- function(in_sample, out_of_sample, mod) {
  # Compute in-sample predictions
  yhat <- predict(mod, newdata=data.frame(Year=in_sample))
  # Display in-sample predictions in green
  text(in_sample, yhat, round(yhat, 2), col="green", cex=0.75)
  # Compute out-of-sample predictions
  yhat <- predict(mod, newdata=data.frame(Year=out_of_sample))
  # Display out-of-sample predictions in red
  text(out_of_sample, yhat, round(yhat, 2), col="red", cex=0.75)
}
```

## Here's some code to display the data and the predictions.

Code for displaying the curve

```{r function3, echo=TRUE}
draw_curve <- function(in_sample, out_of_sample, mod) {
  # Compute in-sample curve
  yhat <- predict(mod, newdata=data.frame(Year=in_sample))
  # Display in-sample predictions in green
  lines(in_sample, yhat, col="green")
  # Compute out-of-sample curve
  yhat <- predict(mod, newdata=data.frame(Year=out_of_sample))
  # Display out-of-sample predictions in red
  lines(out_of_sample, yhat, col="red")
}
```

## Here's some code to display the data and the predictions.

Finally, code to put it all together

```{r function4, echo=TRUE}
draw_complete_plot <- function(n) {
  draw_raw_data()
  mod <- lm(Rate ~ poly(Year, degree=n), data=d1)
  in_sample_points <- d0$Year[-15]
  out_of_sample_points <- d0$Year[15]
  draw_predictions(in_sample_points, out_of_sample_points, mod)
  in_sample_sequence <- seq(d0$Year[1], d0$Year[14], length=100)
  out_of_sample_sequence <- seq(d0$Year[14], d0$Year[15], length=100)
  draw_curve(in_sample_sequence, out_of_sample_sequence, mod)
}
```

## Let's look at a sequence of polynomials.

<div class="columns-2">

```{r fit_1, fig.width=4, fig.height=5}
draw_complete_plot( 1)
```


The linear regression fits reasonably well. The prediction for 1997, the "out-of-sample" prediction, is not too far from 2.3. 

</div>

## Let's look at a sequence of polynomials.

<div class="columns-2">

```{r fit_2, fig.width=4, fig.height=5}
draw_complete_plot( 2)
```

The quadratic regression looks a little bit better. In particular, the "out-of-sample" prediction looks to be improved.

</div>

## Let's look at a sequence of polynomials.

<div class="columns-2">

```{r fit_3, fig.width=4, fig.height=5}
draw_complete_plot( 3)
```

This is a bit disturbing. The cubic regression flattens things out at the end, making for a slightly worse "out-of-sample" prediction. The reason it flattens out is that the last four murder rates: `r paste(d0$Rate[11:14], collapse=", ")` are somewhat flat and your cubic polynomial is trying hard (maybe a bit too hard) to model this late flattening trend. 

</div>
